Implementing \& Evaluating Space Efficient Algorithms for Detecting Large Neighbourhoods in Graph Streams

### 2020-04-19 paper/src/paper.tex
Insertion-Only write up mostly done

\usepackage[ruled, vlined, linesnumbered]{algorithm2e}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{changepage}
\usepackage{fancyhdr} % Header

@@

\renewcommand{\headrulewidth}{0pt}
\newcommand{\ie}{\textit{i.e.} }
\newcommand{\nats}{\mathbb{N} }
\newcommand{\horizontalline}{\newline\vspace{.3cm}\hfill\makebox[.5\linewidth]{\rule{.5\textwidth}{0.4pt}}\hfill\vspace{.1cm}}
% title
\title{Implementing \& Evaluating Space Efficient Algoirthms for Detecting Large Neighbourhoods in Graph Streams}
\author{Dom Hutchinson}
\date{\today}
\maketitle

@@

\chapter{Background}

\section{Motivation}
\par For most algorithms time efficiency is viewed with primary importance, rather than space efficiency. However if the memory used for an algorithm exceeds the size of the computer's RAM then virtual memory would have be used. Virtual memory has a much longer read/write than RAM meaning that the algorithm's run-time will be much greater in this scenario. Thus for algorithms which are designed to work with large data sets space efficiency is more important.
\par In the early 2000s companies began to realise that data could be used as a commodity, and with this the amount of data being collected has surged in the years since. In this same time period we have nots seen the same surge in the amount of RAM typical computers have. Thus people wishing to capitalise on this increase in available data need to give greather thought to the space efficiency of the algorithms they are using. Otherwise run-times could become unusuable.
\par Social networks are good sources of data about people. This data can leveraged in many ways, notably within advertising. Behaviourly targeted adverts have been shown to have more than twice the click through rate of standard online advertiesments \cite{targetedAds}. The last decade has seen developments in the space of targeted advertising, one of which is the rise of social media influencers. These are people who are deemed to have a large online following, which they are able to influence. Advertisers will often approach these influencers with deals by which the influencer will promote a product to their audience, hopefully driving sales.
\par The question becomes, ``Who is an influencer \& how do we find them?''. This question can be generalised as the $\mathtt{Neighborhood\ Detection\ Problem}$. Which is the problem I shall discuss in this paper.
%TODO
Although in my description of the Neighbourhood Detection Problem I validate its relevance by referring to find social media influencers, it is applicable to many other scenarios.
\begin{itemize}
	\item[-] TODO
\end{itemize}
\section{Neighbourhood Detection Problem}
%TODO Describe Graph streams, insertion-only & insertion-deletion
\par Graph streams are a dynamic way of representing graphs. They are simply a list of instructions for how to construct a graph and in this paper I shall discuss two types: insertion-only streams \& insertion-deletion streams. There are other types which allow for updating vertex values, among other properties, but they are not relevant to the problem we are considering in this paper.
\par Insertion-only streams are a list of pairs of vertices, representing endpoints of edges. Each time we read an entry from an insertion-only stream we are adding the edge it describes to the graph. We apply a limitation that no duplicate edegs appear in an insertion-only stream and thus the order of the stream has no affect on the validity of the algorithms which shall be discussed.
\par Entries in an insertion-deletion list specify a pair of vertices, representing an edge, and a boolean which states whether this is an insertion or deletion entry. If it is an insertion entry then we are adding the described edge to the graph; if it is a deletion entry then we are removing it. Again we add a limitation that no edges are repeated \& that we can not recieve a request to delete an edge from the graph which does not currently exist in the graph. This does however mean that the order of an insertion-deletion is important as the deletion of an edge must come after it.
\par Despite these limitations it is still possible to construct almost any type of graph from a graph stream. Suppose you wish to analyse which devices in a network are making the most request. Here vertices represent devices \& edges represent requests. In this scenario you would want to allow for multiple edges (say $n>2)$ between the same pair of devices (say $\{x,y\}$). One solution would be to add $n$ new vertices $\{v_{x,y,1},\dots,v_{x,y,n}\}$ each with a label stating they are just indetermiate nodes and not full devices. Creating edges from $x$ to each of $\{v_{x,y,1},\dots,v_{x,y,n}\}$ \& from $y$ to each of $\{v_{x,y,1},\dots,v_{x,y,n}\}$ allows for an interpretation of multiple edges.
\par If a graph stream is ordered by time (which is common in real life applications) then it represents a dynamic-evolution of the graph over time. Allow the user to roll back the graph should they wish to make assessments during particular time preiods, rather than just assess the final graph.
\par The algorithms which shall be discussed in this paper are designed to work with just a single pass of a graph stream. This is not only time efficient as it does not require multiple reads of the stream, but means that should more information be gained after the algorithm is run it is easy to incorporate it into algorithm if we store the final state of the data structures used. Single pass algorithms mean we don't necessarily have to store the graph stream, we could simply send entries from the system being modelled straight into the algorithm, without being stored in a file. This is useful when a graph stream potentially could take up multiple terra-bytes, but is unnecessary in most applications.

>>>>

\usepackage[margin=1in, paperwidth=8.3in, paperheight=11.7in]{geometry}

\usepackage[ruled, vlined, linesnumbered]{algorithm2e}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{changepage}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{fancyhdr} % Header
\usepackage{graphicx}

@@

\renewcommand{\headrulewidth}{0pt}
\newcommand{\ie}{\textit{i.e.} }
\newcommand{\nats}{\mathbb{N} }
\newcommand{\horizontalline}{\newline\vspace{.3cm}\hfill\makebox[.5\linewidth]{\rule{.5\textwidth}{0.4pt}}\hfill\vspace{.05cm}}

\renewcommand\thetable{\arabic{table}} % figure numbering strategy
\renewcommand\thefigure{\arabic{figure}} % figure numbering strategy

\pagenumbering{roman}

@@

% Algorithm environment
\SetArgSty{textnormal}
\DontPrintSemicolon

% Title
\title{Implementing \& Evaluating Space Efficient Algorithms for Detecting Large Neighbourhoods in Graph Streams}
\author{Dom Hutchinson}
\date{\today}
\maketitle
\newpage

\tableofcontents
\newpage

\renewcommand\thechapter{\Roman{chapter}}
\renewcommand\thesection{\thechapter.\roman{section}}
\setcounter{chapter}{1}
\chapter*{Summary/Abstract}
\addcontentsline{toc}{chapter}{Summary/Abstract}

\section{Background}

%\begin{itemize}
%	\item High Degree Detection.
%	\item Large Neighbourhood Detection.
%\end{itemize}

%TODO testing space usage is important as it suggests what will happen when larger graphs are used (the longer we can keep space requirements below RAM space the better).

 % TODO is it ok to define streams & such later
Graphs are a popular way of representing relationships between objects. They are particularly popular within the field of data science \& with the recent movement of \textit{Big Data} it is becoming common to want to deal with large graphs. Moreover, it is common to want to use graphs which cannot be loaded fully onto a computer's RAM and instead require the use of virtual memory. Having to use virtual memory drastically increases request time for data \& thus negatively affects the runtime of algorithms which manipulate these large graphs. Often, in practice, the fastest algorithms for analysing graphs are those which are more space efficient. %rather than those that, in theory, have lower run-time complexity
\par Graph streams have been introduced to reduce the space required to store a graph since they require $O(|E|)$ space, rather than $O(|V|^2)$ space which a traditional matrix representation requires. % TODO define |E| & |V| as edge and vertex sets
Since the number of edges in a graph is cannot be greater than, and is usually much lower than, the square of the number of vertices this is a big improvement. However many graphs still exist which would overflow RAM when loaded in full when in stream form. This is solved by reading the stream one entry at a time as this requires $O(1)$ space. Thus the most space-efficient algorithms for analysing graphs are streaming-algorithms which read instructions sequentially.
\par Graph streams can be used to show the evolution of a graph by insertion \& deleting edges over some time period, in these implementations it is possible for a graph stream to require more space than a matrix representation but since only one instruction is every loaded into the RAM at one time this does not affect the total memory used by a streaming-algorithm. %TODO this might seem out of place (too much detail too early) but I want to make the point that O(|E|) can be greater than O(|V^2|) and this doesn't matter
% TODO perhaps mention that graph streams can show the evolution of a graph (in this case they may have more entries than |V|^2 but doesn't matter since only one entry is every read at a time).
% TODO why single pass??
\horizontalline
\par High degree detection is a common and trivial problem in graph theory. The \texttt{High Degree Detection Problem} entails finding a vertex in a graph with sufficiently high degree. This can be trivially solved for graph streams by keeping counters for the number of edges connected to each vertex and updating the counters after each instruction from the graph stream. % This requires $O(|V|)$ space
\par Although it is useful to identify influential vertices in a graph, it is hard to fully utilise this knowledge without additional information about the neighbourhood of these vertices. Knowing the neighbourhood of a vertex, or at least a subset of it, allows you to make inferences about that node. This leads to the \texttt{Neighbourhood Detection Problem} which is formally stated as

\vspace{.3cm}\begin{adjustwidth}{.3cm}{}\fbox{\parbox{\textwidth}{
\textbf{Problem 1} \texttt{Neighbourhood Detection}.\\
Let $G=(A\cup B,E)$ be a bi-partite graph with vertex sets $A,B$, where $|A|=n$ and $|B|=\text{poly }n$, and edge-set $E$.\\
In $\texttt{Neighbourhood Detection}(G,d,c)$ we are tasked with outputting a vertex from $A$ with at least $d/c$ of its neighbours in $B$. We can assume that $G$ contains at least one node of degree $d$.\\
Here $d\in\nats$ is a threshold parameter \& $c\in(0,1]$ is an approximation parameter.
}}\end{adjustwidth}\vspace{.3cm}

The \texttt{Neighbourhood Detection Problem} is defined for bi-partite graphs in order to generalise the problem. Note that, if we were to define the vertex sets $A,B$ to be equivalent then then the problem holds for every type of graph.
\par Again, the \texttt{Neighbourhood Detection Problem} can be solved trivially by storing the neighbourhoods of every vertex in the graph \& updating them after each instruction from the graph stream. This requires $O(|E|)$ space and can very quickly lead to overflowing the RAM which is undesirable since it means slower run-times for large graphs.

\section{Motivating Applications} %TODO I don't really like this (partly because it is in list form), moreover it isn't really connected in
% NOTE maybe it is best that it is fairly separate from the rest so that it can be skipped over.
%\begin{itemize}
%	\item Real world occurrences \& uses.
%\end{itemize}

There are several real-world applications of the \texttt{Neighbourhood Detection Problem} which motivate investigating it. These are situations where simply knowing highly connected elements of a network is not sufficient.
\begin{itemize}
	\item[-] Given a list of connections within a social network, solving the problem would allow you to not only find \textit{`social influencers'} but those they are connect to too. This allows you to verify that they are indeed a \textit{`social influencers'}, and not someone connected to a lot of bot accounts, and to assess what demographics they have influence over.\\
	For a marketing agency this would mean they don't waste money on people who don't actually have any influence to sell a product, nor on people who have the wrong demographics for a certain product.
	\item[-] Given a list of passengers on public transport you could find people who have come into contact with a lot of people. This is particular relevant in the wake of the COVID-19 outbreak as it could be used to identify people who are likely to have come into contact with infected people. %TODO not a fan of these, might just need fleshing out
	\item[-] Given a list of receipts in a shop you could find the items which are popular, along with which items they are commonly bought with. This information could be used to inform how shelves are stacked and which promotions are run by the store, in order to encourage sales.
	\item[-] Given a log of traffic within a network you could identify which resources are being accessed most often, and by whom. This can be used to inform which resources should be upgraded \& which connections should be prioritised.
\end{itemize}

\section{Structure} %TODO after most the write up
In this paper I am going to implement the two algorithms proposed by Dr Christian Konrad in \cite{orig} for solving the \texttt{Neighbourhood Detection Problem} for graphs streams in a space-efficient manner. I will use this implementation to assess how practical these algorithms are and now to maximise their practicalities.
\par I shall structure this investigation as follows

\begin{itemize}
	\item Insertion-Only Streams
	\begin{itemize}
		\item Layout theory
		\item Discuss features that need to be/can be adjusted in the implementation.
		\item Implement - mentioning any other problems.
		\item Evaluate \& test different combinations of parameters to optimise the practical uses of the algorithm.
		\item Discuss limitations of the implementation. (ie What graphs work)
	\end{itemize}
	\item Repeat for Insertion Deletion.
\end{itemize}

\section{Objectives}
The objectives of this project are
\begin{itemize}
	\item Implement Christian's algorithms and discuss any considerations that need to be taken in doing so.
	\item Tune parameters to make algorithms as practical as possible.
	\item How practical are the algorithms.
\end{itemize}

\renewcommand\thechapter{\arabic{chapter}}
\renewcommand\thesection{\thechapter.\arabic{section}}
\setcounter{chapter}{0}
\chapter{Preliminaries}

\setcounter{page}{1} % restart page counter
\pagenumbering{arabic}

\section{Definitions}

%\begin{itemize}
%	\item Graph
% \item Unweighted, undirected
%	\item Bipartite Graph (Why this wasn't really applied in practice, A=B)
%	\item Graph Stream (Insertion-Only, Insertion-Deletion)
%	\item Neighbourhood, degree & sparsity of a vertex
%	\item Streaming Algorithm
%	\item $1-\varepsilon$ Approximation Algorithm
%	\item TODO Space efficiency??
%\end{itemize}

A \textit{Graph} is a data structure used to represent pairwise relationships between objects. A Graph is defined to have a vertex set, which holds its objects, and an edge-set, which holds the relationship between these objects. Graphs are traditionally visualised with circles for each vertex and lines between vertices that share an edge. There are variations on graphs where the edges are weighted and/or directional. In this project we are only required to deal with undirected, unweighted graphs.
\par A \textit{Bipartite Graph} is a graph whose vertex set can be partitioned into two distinct sets in such a way that all edges in its edge-set run between these partitions \& not within them. % Formally, a graph $G=(V,E)$ where $\exists\ A,B\subset V$ st $V=A\cup B,\ A\cap B=\emptyset$ and $\forall\ (u,v)\in E,\ u\in A\ \&\ v\in B$.
\par The \textit{Neighbourhood} of a vertex in a graph is the set of vertices which it shares an edge with. The \textit{Degree} of a vertex is the size of its neighbourhood. A vertex is $s$ \textit{Sparse} if its degree is less than, or equal to, $s$.
\par A \textit{Graph Stream} is a technique for storing graphs. A graph stream is made up of a series of instructions which are read sequentially and describe how to construct the graph. These instructions describe modifications to the edges of the graph. In this project we are only concerned with two types of graph stream.
\begin{itemize}
	\item \textit{Insertion-Only Streams} where each instruction adds a new edge to the graph.
	\item \textit{Insertion-Deletion Streams} where each instruction either adds a new edge to the graph or removes an existing edge from the graph.
\end{itemize}
Neither of these allow for duplicates of the same edge in the graph, at the same time. Insertion-Deletion streams are used for graphs whose edge-set is dynamic over time, consider a social network where users are able to follow \& un-follow each other. There are other versions of graph streams which all for changes to the weights of edges but these are out of scope for this project. It is noteworthy that insertion-only streams equivalent to insertion-deletion streams which have no deletions. This means that they can be implemented in exactly the same way. % TODO maybe state that these are equivalent to the above if we treat 0 weight to be no edge & anything else to be an edge %TODO maybe mention when order is important (or can represent time)
\par \textit{Streaming Algorithms} are algorithms which are designed to take a stream of sequential instructions as their input. Streaming algorithms can be very space-efficient since only one instruction is read at a time, requiring constant space. Typically streaming algorithms aim to only require a single pass of the input stream so that you can extend the input stream without having to restart the algorithm. Single-pass streaming algorithms are online algorithms. %TODO make the mention of online algorithms more succinct
\par An \textit{$\alpha$-Approximation Algorithm} is an algorithm which returns a result within an $\alpha$ factor to the optimal solution to the problem. Algorithms which solve the \texttt{Neighbourhood Detection Problem} are $\frac1c$-approximation algorithms since they only return $d/c$ of the neighbours of a vertex which we know to have at least $d$ neighbours.

\section{Technologies}
%\begin{itemize}
%	\item C++
%	\begin{itemize}
%		\item Low level
%		\item Memory manipulation
%		\item Faster
%		\item Perhaps something about why not java or python etc.
%	\end{itemize}
%	\item Standard library online
%\end{itemize}
For the implementations in this project I chose to use C++. I initially considered using either python or C++, as they work in the object-orientated paradigm and I am proficient in them. I chose to go with C++ (specifically \texttt{C++11}) as it allows for more control with memory management and typically faster run-times. For this project the memory management is more important as I am looking to evaluate the space-efficiency of algorithms and being able to control their memory allocation means I can identify each time a change occurs to the amount of memory being used. The faster run-times is important when evaluating the practicalities of using these algorithms, as longer run-times make algorithms less practical.
\par I limited myself to using the \textit{C++ Standard Library} (\texttt{std}) as the underlying implementations of this library are well document. This was particularly important for the abstract data types I used as I could check what the underlying implementations and adjust my evaluation accordingly. When implementing randomness I used the \texttt{<random>} module of \texttt{std} and seeded the generators with the current time so that each run would have a different generator.

\section{Evaluation Techniques}

% TODO expectations
%\begin{itemize}
%	\item What features are tested (time \& space)? How?
%	\item Against naïve algorithm.
%	\item Insertion-Only v Insertion-Deletion algorithm for insertion-only streams.
%\end{itemize}

%TODO should this be in the future tense (eg in this paper i will)
%TODO maybe keep this more vague and move the details to their respective sections
%TODO talk about verifying my results
%TODO typical test = vary c between [3,20]. set d=max degree

When evaluating these algorithms there were three main parameters which needed to be tested for: success-rate; time-taken; and, space-used. These are all interconnected and come with their own trade-offs. Typically, the changes to space-used \& time-taken when different strategies are implemented occur in the same direction but at different rates. Whereas, you expect success-rate to go in the opposite direction as the algorithm is using fewer resources.
For these particular algorithms the success rate always remained high but for some setups it takes a lot longer to find a solution. %TODO do I means this??
\par The evaluation process of each algorithm can be broken into two stages. First, I evaluated possible efficiencies to the underlying implementation. The changes tested at this stage did not affect the success-rate of the algorithms but did, in some cases, have a drastic effect on the amount of resources used by each algorithm. Second, I evaluated different set-ups for the parameters of the algorithm. This stage involved a lot more \textit{tuning} of values \& strategies as these affected the logic of the algorithm resulting in trade-offs between resource requirements \& success-rate. The changes in the first were a lot more general \& there was little variation in the results seen for different graph types, whereas in the later stage I had to consider different graph types.
\par I used a global variable to count the space used by different components of the algorithms (as well as the algorithm as a whole) and a timer to evaluate the run-time. For each variation I ran the tests multiple times, recording the results of each run. Typically for testing I set $d$ to be the maximum degree of the graph as these algorithms are typically used to investigate the nodes of greatest degrees in a graph, and I would then vary the approximation factor $c$ as seem reasonable for the given graph.

\subsection{Generating Graphs}

%\begin{itemize}
%	\item How graphs streams are represented.
%	\item Different graph types.
%	\item Generating graphs
%	\item Hard to have wide variety of node degrees.
%	\item Acquired from SNAP.
% \item Graph pre-processing (.vertices file).
%	\item Programs to verify success (ie returned neighbourhood is valid).
%\end{itemize}
For this project I required a number of graphs to evaluate my programs with. By the nature of the algorithms these graphs had to be undirected \& unweighted bi-partite graphs. The bi-partite requirement is unimportant as any graph can be treated as bi-partite by making the vertex set partitions copies of each other.
\par I implemented programs to generate graphs with specific properties which I wanted to test, namely graphs with high disparity in the degree of each node (such as star graphs) and graphs with no disparity between degrees (such as complete graphs). % TODO more justification
I implemented the following programs.
\begin{itemize}
	\item \texttt{completeGraph.cpp} - Generates a graph stream for a complete graph.
	\item \texttt{randomGraph.cpp} - Uses a Bernoulli distribution to determine whether two vertices share an edge in an insertion-only stream, and a poisson distribution to determine how many insertions \& deletions occur between each pair of vertices.
	\item \texttt{starGraph.cpp} - Generates a graph stream for a star graph.
\end{itemize}
All these programs take parameters for the number of vertices and \texttt{randomGraph.cpp} takes parameters for the proportion of edges. Additionally these, programs are able to generate insertion-only or insertion-deletion streams but the final graphs defined by the streams generated by \texttt{completeGraph.cpp} \& \texttt{starGraph.cpp} are always of the appropriate shape. %TODO reword this
%TODO I found it hard to have a wide variety of node degrees
\par I wanted some graphs generated from the real-world so that I could test the algorithms on the sort of data sets they are most likely to be used on. I found a collection of large graph streams, from different applications, in the \texttt{Stanford Network Analysis Project} (\texttt{SNAP}) datasets collection \cite{SNAP}. %TODO I chose social network ones??
The vertices in the instructions of these graph streams were not ordered meaning that these graphs are not strictly bi-partite. % TODO better explanation?
This meant I needed to change my implementations slightly such that both ends of the edge are treated as $A$-vertices. This is equivalent to having a bi-partite graph where the $A$ \& $B$ vertex sets are copies of each other, which is perfectly fine.
\par The graphs from \texttt{SNAP} were all insertion-only streams so I wrote a program which would generate an insertion-deletion stream from them by adding a defined proportion of deletion instructions throughout the stream. This program maintained the property that no edge was deleted before it was inserted.
\par I wrote \texttt{utility.cpp} to perform a number of tasks to these graphs. Most notably, it can produce a file which lists the final degree of every vertex in the graph stream and it can verify whether a given set is a subset of the neighbourhood of a given vertex. I used the later function to verify the outputs generated by implementation of Dr Konrad's algorithms.
%TODO table of used graphs

%TODO confirm values for gplus_large
\begin{center}
	\begin{table}[h]
		\caption{Details of graphs used during evaluation}
		\tiny
		\begin{tabular}{|l|l|l|l|l|l|l|l|}
			\hline
			\textbf{Name}&\textbf{Source}&\textbf{Type}&\textbf{\# Vertices}&\textbf{\# Edges}&\textbf{\# Instructions}&\textbf{Max Degree}&\textbf{File Size}\\
			\hline
			$\mathtt{facebook\_small}$&SNAP&Insertion-Only&52&292&-&36&3 KB\\
			$\mathtt{facebook\_small\_deletion}$&Generated&Insertion-Deletion&52&258&326&33&5 KB\\
			$\mathtt{facebook}$&SNAP&Insertion-Only&747&60,050&-&586&587 KB\\
			$\mathtt{facebook\_deletion}$&Generaetd&Insertion-Deletion&747&53,457&66,643&522&846 KB\\
			$\mathtt{gplus}$&SNAP&Insertion-Only&12,417&1,179,613&-&5,948&53 MB\\
			$\mathtt{gplus\_deletion}$&Generated&Insertion-Deletion&12,417&1,049,309&1,309,917&4,998&60 MB\\
			$\mathtt{gplus\_large}$&SNAP&Insertion-Only&102,100&30,238,035&-&104,947&1.3 GB\\
			$\mathtt{1000vertices}$&Generated&Insertion-Only&1,000&246,676&-&984&2 MB\\
			$\mathtt{doublepoisson05ID}$&Generated&Insertion-Deletion&10,000&8,639,534&24,452,736&4,977&305 MB\\
			$\mathtt{1000star}$&Generated&Insertion-Only&1,000&999&-&999&9 KB\\
			$\mathtt{1000starID}$&Generated&Insertion-Deletion&1,000&999&1,023&999&11 KB\\
			$\mathtt{1000complete}$&Generated&Insertion-Only&1,000&499,500&-&999&4 MB\\
			$\mathtt{1000completeID}$&Generated&Insertion-Deletion&1,000&499,500&509,524&999&5 MB\\
			\hline
		\end{tabular}
	\end{table}
\end{center}

\chapter{Insertion-Only Streams}

\section{Degree-Based Reservoir Sampling}

\texttt{Degree-Based Reservoir Sampling} is a technique for uniformly sampling a subset from the set of vertices whose degrees are greater than a specified minimum bound, while storing part of the neighbourhood for each of the sampled nodes. This sample is known as the \textit{reservoir}.

\par \texttt{Degree-Based Reservoir Sampling} takes three parameters: the lower bound for the degree of vertices at which to start sampling, $d_1\in\nats$; the maximum neighbourhood size to store for each vertex, $d_2\in\nats$; and the maximum reservoir size, $s\in\nats$. \texttt{Degree-Based Reservoir Sampling} succeeds if it finds a neighbourhood of size $d_2$.

%\par Degree-based reservoir sampling \underline{only} works for insertion-only graph streams.

\par Three data stores are used in \texttt{Degree-Based Reservoir Sampling}: a map of the degree of every node in the graph; a set for the reservoir; and a set for edges incident to the vertices in the reservoir.

\par The reservoir has an invariant that at any point in time it contains a uniform sample of the vertices whose degrees are known to be greater than $d_1$ at that time. This invariant is maintained by controlling how vertices are added to the reservoir. The first time the degree counter for a vertex $v$ surpasses $d_1$ we consider whether to add $v$ to the \textit{reservoir}. There are two cases
\begin{itemize}
	\item If the reservoir is \underline{not} full (ie its size is less than $s$) then we add $v$ to the reservoir.
	\item Otherwise, if the reservoir is already full then we use a Bernoulli random variable with probability $p=\frac{x}s$, where $x$ is the size of the reservoir. If this random variable succeeds then we pick, uniformly at random, a vertex $u$ currently in the reservoir and replace it with $v$. We remove all the edges in the edge set which are incident to $u$ and no other vertices in the reservoir.
\end{itemize}
If $v$ is not added to the reservoir at this time, it never will be in the future. Note that in the case that the reservoir is full \& $v$ is inserted into it, $v$ is replacing an element that was in the reservoir and any progress which has been made towards finding a neighbourhood for that vertex is annulled in doing so. The amount of progress is not taken into account \& thus if there is a high turn-over of elements in a reservoir then less progress is made towards finding a neighbourhood for each of them. This is something I will look into during the evaluation.

\par Whenever an edge is encounter which is in incident to a vertex in the reservoir then we add it to the set of edges. The sampler runs until the graph stream is exhausted at which point it returns, uniformly at random, one of the neighbourhoods of sufficient size.

\par \texttt{Degree-Based Reservoir Sampling} only works on insertion-only streams as it does not account for when the degree of a vertex falls below the threshold, after being sampled. This means the reservoir is not necessarily a uniform random sample of the vertices with degree $\geq d_1$. %TODO is this sufficient

\par For this project we are interested in how \texttt{Degree-Based Reservoir Sampling} can be implemented for bi-partite graphs, specifically we want to sample from the $A$-vertices of the graph. For a bi-partite graph with $n$ $A$-vertices \texttt{Degree-Based Reservoir Sampling} requires $O(n\log n+sd_2\log n)$ space, assuming $O(\log n)$ space is required to store an edge or a vertex, since it stores a degree counter for every $A$-vertex and at most $d_2$ edges for each of the $s$ vertices in the reservoir. \textbf{Algorithm 1} outlines pseudocode for \texttt{Degree-Based Reservoir Sampling} on a bi-partite graph.

\begin{algorithm}
	\caption{\texttt{Degree-Based Resevoir Sampling}$(d_1,d_2,s)$}
	\SetKwInOut{Require}{require}
	\Require{Lower degree bound $d_1\in\nats$, Upper neighbourhood bound $d_2\in\nats$, Reservoir size $s\in\nats$, Insertion-only stream $\{(a_0,b_0),\dots,(a_n,b_n)\}$}
	$D\leftarrow\{\}$ \tcp*{Degree counter}
	$R\leftarrow\{\}$ \tcp*{Reservoir}
	$S\leftarrow\{\}$ \tcp*{Collected edges}
	$x\leftarrow0$\tcp*{\# nodes of degree$\geq d_1$}
	\For{$i\in[0,n]$} {
		$D[a_i]\leftarrow D[a_i]+1$\tcp*{Increment degree counter}
		\If{$D[a_i]\equiv d_1$}{
			\tcp{Consider inserting $a_i$ to reservoir}
			$x\leftarrow x+1$\\
			\If{$|R|<s$} {
			\tcp{Reservoir is not full}
			$R\leftarrow R\cup\{a_i\}$
			} \Else {
				\tcp{Reservoir is full}
				\If{\texttt{Bernoulli}$\left(\frac{s}x\right)$} {
					Let $a'$ be a uniform random element of $R$\tcp*{Element to replace}
					Delete edges in $S$ incident to $a'$\\
					$R\leftarrow(R\backslash\{a'\})\cup\{a\}$\tcp*{Swap $a_i$ and $a'$}
				}
			}
		}
		\If{$a_i\in R$}{
			$S\leftarrow S\cup(a_i,b_i)$\tcp*{Store edge}
			\If{$D[a_i]\equiv d_1+d_2$} {
				\Return{edges in $S$ incident to $a_i$\tcp*{Success}}
			}
		}
	}
	\Return{FAIL\tcp*{No sufficient neighbourhoods found}}
\end{algorithm}

\section{Proposed Algorithm}

%\begin{itemize}
%	\item Assumptions: no duplicate edges (can mitigate cases when this would be required by being smart)
%	\item Insertion stream neighbourhoods cannot be diminished over time.
%	\item Possibility of failure (missed edges) \& mitigation
%	\begin{itemize}
%		\item Theorem 3.2 (probs not proof)
%	\end{itemize}
%	\item Theoretical space requirements
%\end{itemize}

The \texttt{Neighbourhood-Detection Problem} can be solved using \texttt{degree-based reservoir sampling} sampling by setting $d_2=\frac{d}c$. The question remains as to what to set $d_1$ and $s$ to in order to get a high probability of success. In \cite{orig} it is proven that setting $s=\lceil\ln(n)\cdot n^{\frac1c}\rceil$ and using $c$ \texttt{degree-based reservoir samplers} each with a different lower bound, incremented from $\frac{d}c$ to $d$ stepping by $\frac{d}c$ each time, results in a high probability of success. The algorithm suggested by Dr. Konrad runs the samplers in this way, with $s$ set to $\lceil\ln(n)\cdot n^{\frac1c}\rceil$. The proposed algorithm returns, uniformly at random, any of the neighbourhoods return by a successful sampling. %TODO proof? Say it's insertion-only
\par Each of the $c$ samplers requires $O(n\log n+sd_2\log n)$ space. Since the degree map can be shared between samplers the total space is $O(n\log n+c\cdot sd_2\log n)=O(n\log n+c\lceil\log(n)n^{\frac1c}\rceil\frac{d}c\log n)=O(n\log n+n^{\frac1c}d\log^2n)$. %TODO with algorithm parameter values
The proposed algorithm does require you to know the number of vertices in the graph stream before running the algorithm, or at least the log of it. If this is unknown then it is quick to run through the whole stream \& build a set of vertices in the graph. This requires $O(n\log n)$ space, assuming $O(\log n)$ space is required to store a vertex. In many real-world scenarios this number (or a good approximation) will be known due to it being important for other tasks.
\par \textbf{Algorithm 2} is the algorithm proposed in \cite{orig} for solving the \texttt{Neighbourhood-Detection Problem} for insertion-only streams.\\

\begin{algorithm}
	\caption{One-pass $c$-Approximation Insertion-Only Streaming Algorithm for $\mathtt{Neighbourhood\ Detection}$}
	\SetKwInOut{Require}{require}
	\Require{Space $s$, degree bound $d$.}
	$s\leftarrow\lceil\log(n)\cdot n^{\frac1c}\rceil$\\
	\For{$i\in[0,c-1]$\text{\textbf{ in parallel}}} {
		$(a_i,S_i)\leftarrow\mathtt{Deg}\mbox{-}\mathtt{Res}\mbox{-}\mathtt{Sampling}\left(\max\left\{1,i\cdot\frac{d}{c}\right\},\frac{d}c,s\right)$
	}
	\Return{\small Uniform random neighbourhood $(a_i,S_i)$ \text{from successful runs}}
\end{algorithm}

\section{Implementation}

I centred my implementation around working for the graph streams from \textit{SNAP} \cite{SNAP} as they came from real-world sources \& thus are very important for testing the real-world practicalities of the proposed algorithm. These graph streams are not formatted in a strictly bi-partite way. I adjusted \texttt{degree-based reservoir sampling} to account for this by by repeating \texttt{lines 6-19} of \textbf{Algorithm 1} after the end of the \texttt{for} loop, with all occurrences of $a_i$ replaced by $b_i$.
%TODO something about implementing in parallel. Nah
\vspace{.3cm}\newline
After an initial implementation I considered two alterations which could be made:
\begin{enumerate}
	\item \textit{Early Termination} - Returning the first neighbourhood of sufficient size, rather than uniformly sample from all which succeed; and,
	\item \textit{Shared Edge-Set} - Having all samplers share an edge-set, rather than each have one each.
\end{enumerate}

\noindent\textbf{Algorithm 2} states to run through the whole stream before sampling uniformly from the successfully returned neighbourhoods, but this is not necessary to fulfil \texttt{neighbourhood detection}. Returning the first neighbourhood of size $\frac{d}c$ is sufficient for solving the problem. Implementing this should reduce run-time as only part of the stream is being used and should reduce space used by the edge-sets in reservoir sampling as no data is stored after the first solution is found. As this change terminates at the first success there is no change to the success rate of the algorithm. One down side is that you will likely get less variety in the returned neighbourhoods as the algorithm will not encounter every vertex which meets the degree requirements, however this is not relevant to the problem.

\par I implemented this change and evaluated the implementation at different values of $c$. Space-requirements more than halved in every instance. For graphs with the least variation in degree distribution space-requirements decreased by more than $97\%$ after early termination was introduced (as there are many vertices with valid neighbourhoods). Introducing early termination meant that run-times reduced as $c$ increased, whereas before run-times increased with $c$. These results are shown in \texttt{Figure 1} \& \texttt{Figure 2}. These are really good results, so I implemented this change going forward. %TODO explain reasons behind these results
%TODO comment about effect on space used by reservoirs vs degree map
%TODO this change means that time taken is now a measure of success as failure is (basically) the only time the whole stream is run through

\textbf{Lemma 2.1} If \textbf{Algorithm 2} terminates after the first sufficient neighbourhood it finds, no vertex occurs in multiple reservoirs at the same time.\\

% TODO a proof by contradiction
\textit{Proof (by contradiction)} Let $v$ be a vertex in the graph stream. Suppose, without loss of generality, that $v$ is a member of the reservoirs of both \texttt{Deg-Res-Sampling}$(\frac{d}{c},\frac{d}{c},s)$ \& \texttt{Deg-Res-Sampling}$(\frac{2d}{c},\frac{d}{c},s)$. This means that $v$ was sampled twice, after encountering the $\frac{d}c^\text{th}$ \underline{and} $\frac{2d}c^\text{th}$ edge incident to $v$ in the stream. After sampling $v$ for the first time, encountered edges which are incident to $v$ are stored. This means that by the time $v$ is sampled for the second time $\frac{2d}c-\frac{d}c=\frac{d}c$ edges incident to $v$ have been stored. This is a sufficient neighbourhood for \textbf{Algorithm 2} to terminate. This is a contradiction since termination occurs before the second sampling occurs.$\hfill\square$

}}\end{adjustwidth}\vspace{.3cm} %TODO refine
